Self-reported LLM usage and outcomes on a data science project: Evidence from a Canadian undergraduate data science course

### 2025-03-14 paper/paper.qmd
small edits

The main variable of interest, `llm_usage` is a categorical variable with three possible values: "Extensive", "Somewhat", "None or minimal". We use "None or minimal" as the reference level and so the results are in relation to that level of self-reported LLM usage. Self-reported GPA, `what_is_your_gpa`, is a number between 1.9 and 4 with one decimal place.

Define $y_i$ as the percentage that student $i$ received on the final paper:

@@

We estimate the model and explore the results using the statistical programming language `R` [@citeR], and the 
`brms` [@brms], and 
`modelsummary` [@modelsummary] 
packages. Our code to adjust for the fact that there were no students who got 0% follows @buerkner and @Heiss2021. Model diagnostics are included in [Appendix -@sec-model-details]. Finally, the piecewise approach is a common one for when there may be something different about the 0s or 1s, compared with the other responses. Another approach for dealing with a situation in which there are exact 0s or 1s is to slightly perturb the data. In our case, there is not much difference between getting 99% on a paper and getting 100%. When we estimated that perturbed data model using Beta regression the results were similar.

@@

LLM usage estimates are in relation to "Extensive" LLM usage. In the base model, the coefficients suggest that less usage of LLMs was associated with lower scores. The second model retains this association, but with slightly smaller coefficients. For instance, the negative impact of “LLM usage: Somewhat” is slightly reduced from -0.35 to -0.29, indicating that part of the effect observed in the base model could be explained by the GPA variable.

>>>>

The main variable of interest, `llm_usage` is a categorical variable with three possible values: "Extensive", "Somewhat", "None or minimal". We use "None or minimal" as the reference level, so the results are in relation to that level of self-reported LLM usage. Self-reported GPA, `what_is_your_gpa`, is a number between 1.9 and 4.0, measured with one decimal place.

Define $y_i$ as the percentage (0-100%) that student $i$ received on the final paper:

@@

We estimate the model and explore the results using the statistical programming language `R` [@citeR],`brms` [@brms], and`modelsummary` [@modelsummary]  packages. Our code to adjust for the fact that there were no students who got 0% follows @buerkner and @Heiss2021. Model diagnostics are included in [Appendix -@sec-model-details].

In estimating models of this type, a piecewise approach is a common one for when there may be something different about the 0s or 1s, compared with the other responses. Another approach for dealing with a situation in which there are exact 0s or 1s is to slightly perturb the data. In our case,however, there is little difference between getting 99% on a paper and getting 100%. When we estimated that perturbed data model using Beta regression the results were similar.

@@

LLM usage estimates are in relation to "None or Minimal" LLM usage. In the base model, the coefficients suggest that less usage of LLMs was associated with lower scores. The second model retains this association, but with slightly smaller coefficients. For instance, the negative impact of “LLM usage: Somewhat” is slightly reduced from -0.35 to -0.29, indicating that part of the effect observed in the base model could be explained by the GPA variable.
